Staged Abstract Interpreters

### 2019-09-09 paper/30_stageconc.tex
polish

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete
interpreter by changing the monad type and refactoring the primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala, and then replay the same steps from the unstaged
counterpart. Finally, we briefly describe the code generation.

@@

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
uses syntactic quotations, LMS distinguishes binding-times based on types. 
LMS provides a type constructor @Rep[T]@, where @T@ can be an arbitrary type,
indicating a value of type @T@ will be known at the future stage.
Operations acting on @Rep[T]@ expressions will be residualized as generated
code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

@@

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as lists and maps.
The framework internally constructs a sea-of-nodes intermediate representation (IR)
for the next-stage program when executing the current stage \cite{DBLP:conf/birthday/Rompf16}.
For convenience, the conversion from expressions of type @Rep[_]@ to their IR
is done by using the implicit design pattern. As we will see later,
implementing staging and code generation support for user-defined classes is
also straightforward.

@@

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate on
staged values.  For brevity, we call them \textit{staged monads}. In the code,
we also use a @Rep@ prefix on the constructors and types to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@
are not staged, and that the monadic computation, such as functions of @R => M[A]@,
are also not staged. Instead, the internal data that these monads operate on are
staged, i.e., @R@ and @A@ in the reader monad, @S@ and @A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a value known at the current stage.
Therefore all invocations of @flatMap@ can be reduced before generating code.
The fact that we only stage data but not monadic computation or monadic
values is the reason that we can peel of the monad stack in the generated code.
Now the type @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not as straightforward for the nondeterminism monad
(@SetT@) in the abstract interpreter, because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. In this section, we explicitly distinguish the two
versions of monadic interfaces. Later, we will instantiate the staged set monad
by manually fusing a fragment of the monad stack.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.
As we mentioned before, we will take more care on how functions and
applications are handled.  One problem here is what should we do for
$\lambda$ terms when staging? We cannot directly create a next-stage
defunctionalized closure for it, because that means we still need to
interpret over these closure at the next stage.  The desired goal is to
eliminate the interpretation overhead by staging, so the right thing is to
compile the lambda term with its environment to a next-stage Scala function. To
do this, we need to recursively call the interpreter function at the current stage,
i.e., the @ev@ argument passed to @close@.  The following code implements the
idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

In @close@, we first create a current-stage function @f@ that takes a
next-stage value and store, and returns staged values. Then we use
@emit_complied_clo@ to denote @f@ to a next-stage Scala function, represented
by the type @Rep[Value]@. Inside of function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also the prepared environment and store.
By doing so, we \textit{collapse} the monadic value of @AnsM@ to grounded
values within staging; therefore in the future stage, there is no monads.

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@. But what we can do with these values of type @Rep@ is limited. 
In fact, @fun@ does not have an intensional application operation we can use
directly. We only know that it is a staged value produced by @close@, and
it represents a next-stage Scala function. Therefore, we can only generate a
next-stage function application, also in Scala. @ap_clo@ produces a monadic
value of type @Ans@, so that we can still use the monadic operation @get_store@
to obtain the latest store. Then @emit_ap_clo@ takes the function and argument
values, and the latest store to generate a current-stage representation of the
future-stage application result, which is of type @Rep[ValSt]@, a pair of
values and stores.  After obtaining the result @vs@, we use @put_store@ to
update the store and reify the result into the current-stage monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

Compared with the unstaged version, we can observe a shift of where we use the
evaluation function @ev@: in the unstaged interpreter, the invocation of @ev@
happens in @ap_clo@, i.e., at the application time; while in the staged
interpreter, we eagerly apply @ev@ when denoting the lambda terms to our
next-stage value domains, i.e., at the value-representation time.

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete
interpreter by changing the monad type and refactoring the primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala, and then replay the same steps as we did for the unstaged
counterpart. Finally, we briefly describe the code generation.

@@

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
uses syntactic quotations, LMS distinguishes binding-times based on types. 
To achieve this, LMS provides a type constructor @Rep[T]@, where @T@ can be an
arbitrary type, indicating that a value of type @T@ will be known at a future
stage.  Operations acting on @Rep[T]@ expressions will be residualized as
generated code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

@@

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and for commonly-used data structures such as lists and
maps.  The framework internally constructs a sea-of-nodes intermediate
representation (IR) for the next-stage program when executing the current stage
program \cite{DBLP:conf/birthday/Rompf16}.
For convenience, the conversion from expressions of type @Rep[_]@ to their IR
is done by using the implicit design pattern. As we will see later,
implementing staging and code generation support for user-defined classes is
also straightforward.

@@

\paragraph{Staged Monads}
We use the same monad stack structure as in the unstaged interpreter: a reader
monad with a state monad. The difference here is that, now the monads operate
on staged values.  For brevity, we call them \textit{staged monads}. In the
code snippet, we also use a @Rep@ prefix on the name of constructors and types
to differentiate them.
But it is important to note that the instances of @ReaderT@/@StateT@ are not
staged, and that the monadic computation, such as functions of @R => M[A]@, are
also not staged. Instead, the internal data that these monads operate on are
staged, i.e., values of @R@ and @A@ in the reader monad, and values of @S@ and
@A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ types are added.
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to the @flatMap@ is also a value known at the current stage.
Therefore, all invocations of @flatMap@ can be reduced before generating code.
The fact that we only stage data but not the monadic computation or monadic
values is the reason that we can peel of the monad stack in the generated code
through staging.  Now the type @AnsM[_]@ is instantiated as the same structure
as before, but using the @Rep@ versions of monad transformers:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not as straightforward for the nondeterminism monad
(@SetT@) in the abstract interpreter, because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. Therefore, any traversals over the set has to be
staged.  In this section, we explicitly distinguish the two versions of monadic
interfaces. Later, we will instantiate the staged set monad by manually fusing
the fragment of the monad stack inside of @SetT@.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.  As we
mentioned before, we will illustrate in detail how functions and applications
are handled.  One problem here is what should we do for $\lambda$ terms when
staging them? We cannot directly create a next-stage defunctionalized closure
for it, because that means we still need to interpret over these closures at the
next stage and the interpretation overhead has not been eliminated. 
The solution is to compile the $\lambda$ term with its environment by calling
the evaluator at the current stage. The result is generating a next-stage
Scala function. The evaluator is passed as @ev@ to the method @close@.
The following code implements the idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

In @close@, we first create a function @f@ that takes a next-stage value and
store, and returns staged values and stores. The function @f@ is a
current-stage and we then use @emit_complied_clo@ to denote it to a next-stage
Scala function.
The method @emit_compiled_clo@ produces current-stage representations of
next-stage function values, i.e., of type @Rep[Value]@.
Inside of function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also with the prepared environment and
store.  By doing so, we have \textit{collapsed} the monadic values of @AnsM@ to
grounded values within staging; therefore in the future stage, there is no
monads.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@. But what we can do with these values of type @Rep@ is limited. 
In fact, @fun@ does not have an intensional application operation we can use
directly. We only know that it is a current-stage representation of functions produced by @close@,
whose next-stage form is a Scala function.
With this fact, we can generate a next-stage function application for it, also in
Scala. The method @emit_ap_clo@ produces a current-stage representation
of a next-stage function application, with necessary arguments such as the latest
store @σ@. Meanwhile, we still want the method @ap_clo@ implemented in monadic
style, so that it can be smoothly connected with other parts of the
interpreter.  To achieve this, we lift the result values and stores from
the future stage into our current-stage monad stack, with a binding @vs@.
Finally, we use @put_store@ to update the store with @vs._2@ and return the
value result @vs._1@ into the current-stage monadic value.

Now, compared with the unstaged interpreter, we observe a key difference of where
we use the evaluation function @ev@.  In the unstaged interpreter, the
invocation of @ev@ happens in @ap_clo@, i.e., at the application time; while in
the staged interpreter, we eagerly apply @ev@ when denoting the lambda terms to
our next-stage value domains, i.e., at the value-representation time.
